---
title: "Homework 3"
author: "Vasuda Kapoor"
output: github_document
date: '2022-10-13'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Library calls
```{r}
library(tidyverse)
library(ggridges)
library(patchwork)
library(devtools)
library(hexbin)
```

# Problem 1

How many aisles are there, and which aisles are the most items ordered from?
```{r}
#devtools::install_github("p8105/p8105.datasets", force = TRUE, INSTALL_opts = c('--no-lock'))
library(p8105.datasets)
data("instacart")
instacart_df = instacart %>%
  janitor::clean_names()

n_aisle = instacart_df %>%
  summarise(
    aisle_dist = n_distinct(aisle))
aisle_order = instacart_df %>%
  group_by(aisle_id) %>%
  distinct(order_id) %>%
  count() %>%
  arrange(desc(n))
```
There are `r n_aisle` aisles. Aisle 24 has the most number of orders. 

Make a plot that shows the number of items ordered in each aisle, limiting this to aisles with more than 10000 items ordered. Arrange aisles sensibly, and organize your plot so others can read it.
```{r}
instacart_df %>% 
  count(aisle) %>% 
  filter(n > 10000) %>% 
  mutate(aisle = fct_reorder(aisle, n)) %>% 
  ggplot(aes(x = aisle, y = n)) + 
  geom_point() + 
  labs(title = "Number of items ordered in each aisle") +
  theme(axis.text.x = element_text(angle = 60, hjust = 1))
```
Make a table showing the three most popular items in each of the aisles “baking ingredients”, “dog food care”, and “packaged vegetables fruits”. Include the number of times each item is ordered in your table.
```{r}
instacart_df %>%
  filter(aisle == "baking ingredients" | aisle == "dog food care" | aisle == "packaged vegetables fruits") %>%
  group_by(aisle) %>%
  count(product_name) %>%
  mutate(rank = min_rank(desc(n))) %>% 
  filter(rank < 4) %>% 
  arrange(desc(n)) %>%
  knitr::kable()

```

Make a table showing the mean hour of the day at which Pink Lady Apples and Coffee Ice Cream are ordered on each day of the week; format this table for human readers (i.e. produce a 2 x 7 table)
```{r}
instacart_df %>%
  filter(product_name %in% c("Pink Lady Apples", "Coffee Ice Cream")) %>%
  select(product_name, order_dow, order_hour_of_day) %>%
  group_by(product_name, order_dow) %>%
  summarise(mean_hour = mean(order_hour_of_day)) %>%
  spread(key = order_dow, value = mean_hour) %>%
  knitr::kable(digits = 2)
```


# Problem 2
##Load, tidy, and otherwise wrangle the data. Your final dataset should include all originally observed variables and values; have useful variable names; include a weekday vs weekend variable; and encode data with reasonable variable classes. Describe the resulting dataset (e.g. what variables exist, how many observations, etc).
```{r}
# Load data and clean names
accel_df = read_csv("data/accel_data.csv") %>%
  janitor::clean_names()

accel_df2 = read_csv("data/accel_data.csv") %>%
  janitor::clean_names()

# pivot the activity values from wide to long and convert minute column to numerical type
accel_df = 
  pivot_longer(
    accel_df, 
    activity_1:activity_1440,
    names_to = "minute",
    values_to = "activity_value",
    names_prefix = "activity_"
  ) %>% 
  mutate(minute = as.numeric(minute))
  

# Add weekday vs weekend variable
accel_df = accel_df %>%
  mutate(weekend = if_else(day == "Saturday" | day == "Sunday", TRUE, FALSE))
```
This dataset contains five weeks of accelerometer data collected on a 63 year-old male. It dataset has `r ncol(accel_df)` variables and `r nrow(accel_df)`. It contains the following variables- week number, day_id, day, minute of the day, activity_value, and weekend. The weekend variable shows whether the day falls on the weekend or not. 

##Traditional analyses of accelerometer data focus on the total activity over the day. Using your tidied dataset, aggregate across minutes to create a total activity variable for each day, and create a table showing these totals. Are any trends apparent?
```{r}
total_activity_tbl = accel_df %>%
  group_by(day, week) %>%
  summarise(total_activity = sum(activity_value)) %>%
  pivot_wider(
    names_from = day,
    values_from = total_activity
  ) %>%
  select(week, Monday, Tuesday, Wednesday, Thursday, Friday, Saturday, Sunday) %>%
  knitr::kable()
total_activity_tbl
```
There don't seem to be any trends here. Although, it is worth noting that the activity values for week 4 and 5 are much lower on the weekends compared to weekdays. 


##Accelerometer data allows the inspection activity over the course of the day. Make a single-panel plot that shows the 24-hour activity time courses for each day and use color to indicate day of the week. Describe in words any patterns or conclusions you can make based on this graph.
```{r}
# x axis: minute 1 to 1440
activity_plot =
  accel_df %>%
  ggplot(aes(x = minute, y = activity_value, color = day)) +
  geom_line(alpha = .4) +
  geom_smooth(se = FALSE)
activity_plot

```
It looks like the activity increases gradually over the course of the day, plateaus around minute 750 and then starts falling down again around minute 1250. 
On Sundays, the activity increases disproportionately around minute 600 and on Fridays, the activity increases disproportionately around minute 1250. 

# Problem 3

##Describe dataset. 
```{r}
library(p8105.datasets)
data("ny_noaa") 
```
The ny_noaa dataset has `r ncol(ny_noaa)` variables and `r nrow(ny_noaa)` observations. This dataset contains weather data for all stations in the state of New York. The variables in this dataset include station ID, date, precipitation, snowfall, snow depth, maximum recorded temperature and minimum recorded temperature. This dataset contains `r sum(is.na(ny_noaa))` rows with missing values. 

##Do some data cleaning. Create separate variables for year, month, and day. Ensure observations for temperature, precipitation, and snowfall are given in reasonable units. 
```{r}
# data cleaning
ny_noaa_tidy = ny_noaa %>%
  janitor::clean_names() %>%
  separate(date, into = c("year", "month", "day"), sep = "-") %>%
  mutate(tmax = as.numeric(tmax),
         tmin = as.numeric(tmin),
         year = as.numeric(year)) %>%
  mutate(tmax = tmax/10,
         tmin = tmin/10)
```

##For snowfall, what are the most commonly observed values? Why?
```{r}
snwfall_stats <- ny_noaa_tidy %>%
  count(snow) %>%
  mutate(rank = min_rank(desc(n))) %>%
  filter(rank < 3) %>%
  arrange(desc(n))
snwfall_stats
```
The most commonly observed values are 0 and missing. This is probably because (1) It does not snow on a majority of the days during the year (2) Snowfall data is not collected for a significant number of days during the year. 


##Make a two-panel plot showing the average max temperature in January and in July in each station across years. Is there any observable / interpretable structure? Any outliers?
```{r}
avgtmax_df <- ny_noaa_tidy %>%
  group_by(id, year, month) %>%
  filter(month == "01" | month == "07") %>%
  summarise(avg_max_temp = mean(tmax, NA.rm = TRUE)) %>%
  filter(!is.na(avg_max_temp)) %>%
  filter(month == "01" | month == "07") %>%
  mutate(month = case_when(month ==  "01" ~ "January",
      month ==  "07" ~ "July"))

plot_avgtmax <- avgtmax_df %>%
  ggplot(aes(year, avg_max_temp, group = id, color = id)) +
  geom_line(alpha = 0.3) +
  theme(legend.position = "none") +
  facet_grid(. ~ month) +
  xlab("Year") +
  ylab("Average Maximum Temperature (C)")
plot_avgtmax
```
- The average max temperatures are overall higher for July than for the month of January.

- The average max temperature seems to follow a cyclical pattern where it goes up for a few years and then comes back down. This is true for both months. 

- There don't seem to be any true outliers. 

##Make a two-panel plot showing (i) tmax vs tmin for the full dataset (note that a scatterplot may not be the best option); and (ii) make a plot showing the distribution of snowfall values greater than 0 and less than 100 separately by year.
```{r}
# (i) Create plot of tmax  (y axis) vs tmin (x axis)
tmax_tmin_plot <- ny_noaa_tidy %>%
  ggplot(aes(tmin, tmax)) +
  geom_hex() 
# (ii) density plot
snwfall_plot <- ny_noaa_tidy %>%
  filter(0 < snow & snow < 100) %>%
  mutate(year = as.factor(year)) %>%
  ggplot(aes(x = snow, y = year)) +
  geom_density_ridges(alpha = .2)
# show plot (i) and (ii) in a two panel plot
tmax_tmin_plot + snwfall_plot
```

