---
title: "Homework 3"
author: "Vasuda Kapoor"
output: github_document
date: '2022-10-13'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Library calls
```{r}
library(tidyverse)
library(ggridges)
library(patchwork)
library(devtools)
library(hexbin)
```

# Problem 1
```{r}
#devtools::install_github("p8105/p8105.datasets", force = TRUE, INSTALL_opts = c('--no-lock'))
library(p8105.datasets)
#instacart_df = data("instacart") %>%
  #janitor::clean_names()

#n_aisle = instacart_df %>%
  #summarise(
    #aisle_dist = n_distinct(aisle))
#n_aisle
```



# Problem 2
##Load, tidy, and otherwise wrangle the data. Your final dataset should include all originally observed variables and values; have useful variable names; include a weekday vs weekend variable; and encode data with reasonable variable classes. Describe the resulting dataset (e.g. what variables exist, how many observations, etc).
```{r}
# Load data and clean names
accel_df = read_csv("data/accel_data.csv") %>%
  janitor::clean_names()

accel_df2 = read_csv("data/accel_data.csv") %>%
  janitor::clean_names()

# pivot the activity values from wide to long and convert minute column to numerical type
accel_df = 
  pivot_longer(
    accel_df, 
    activity_1:activity_1440,
    names_to = "minute",
    values_to = "activity_value",
    names_prefix = "activity_"
  ) %>% 
  mutate(minute = as.numeric(minute))
  

# Add weekday vs weekend variable
accel_df = accel_df %>%
  mutate(weekend = if_else(day == "Saturday" | day == "Sunday", TRUE, FALSE))
```
##Traditional analyses of accelerometer data focus on the total activity over the day. Using your tidied dataset, aggregate across minutes to create a total activity variable for each day, and create a table showing these totals. Are any trends apparent?
```{r}

total_activity_tbl = accel_df %>%
  group_by(day, week) %>%
  summarise(total_activity = sum(activity_value)) %>%
  pivot_wider(
    names_from = day,
    values_from = total_activity
  ) %>%
  select(week, Monday, Tuesday, Wednesday, Thursday, Friday, Saturday, Sunday) %>%
  knitr::kable()
total_activity_tbl
# Are any trends apparent?
# Doesn't seem like there are any trends based on weekday/weekend
#ANSWER THIS!!
```
##Accelerometer data allows the inspection activity over the course of the day. Make a single-panel plot that shows the 24-hour activity time courses for each day and use color to indicate day of the week. Describe in words any patterns or conclusions you can make based on this graph.
```{r}
# x axis: minute 1 to 1440
activity_plot =
  accel_df %>%
  ggplot(aes(x = minute, y = activity_value, color = day)) +
  geom_line(alpha = .4) +
  geom_smooth(se = FALSE)
activity_plot

```
It looks like the activity increases gradually over the course of the day, plateaus around minute 750 and then starts falling down again around minute 1250. 
On Sundays, the activity increases disproportionately around minute 600 and on Fridays, the activity increases disproportionately around minute 1250. 

# Problem 3

##Describe dataset. 
```{r}
library(p8105.datasets)
data("ny_noaa") 
# DESCRIBE!!!
```

##Do some data cleaning. Create separate variables for year, month, and day. Ensure observations for temperature, precipitation, and snowfall are given in reasonable units. 
```{r}
ny_noaa_tidy = ny_noaa %>%
  janitor::clean_names() %>%
  separate(date, into = c("year", "month", "day"), sep = "-") %>%
  mutate(tmax = as.numeric(tmax),
         tmin = as.numeric(tmin),
         year = as.numeric(year)) %>%
  mutate(tmax = tmax/10,
         tmin = tmin/10)
```

##For snowfall, what are the most commonly observed values? Why?
```{r}
snwfall_stats <- ny_noaa_tidy %>%
  group_by(snow) %>%
  summarise(
    count = n()
  ) %>%
  arrange(desc(count))
snwfall_stats
```
The most commonly observed values are 0 and missing. This is probably because (1) It does not snow on a majority of the days during the year (2) Snowfall data is not collected for a significant number of days during the day. 

##Make a two-panel plot showing the average max temperature in January and in July in each station across years. Is there any observable / interpretable structure? Any outliers?
```{r}
avgtmax_df <- ny_noaa_tidy %>%
  group_by(id, year, month) %>%
  filter(month == "01" | month == "07") %>%
  summarise(avg_max_temp = mean(tmax, NA.rm = TRUE)) %>%
  filter(!is.na(avg_max_temp)) %>%
  filter(month == "01" | month == "07") %>%
  mutate(month = case_when(month ==  "01" ~ "January",
      month ==  "07" ~ "July"))

plot_avgtmax <- avgtmax_df %>%
  ggplot(aes(year, avg_max_temp, group = id, color = id)) +
  geom_line(alpha = 0.3) +
  theme(legend.position = "none") +
  facet_grid(. ~ month)
plot_avgtmax
```

## Make a two-panel plot showing (i) tmax vs tmin for the full dataset (note that a scatterplot may not be the best option); and (ii) make a plot showing the distribution of snowfall values greater than 0 and less than 100 separately by year.
```{r}
# (i) Create plot of tmax  (y axis) vs tmin (x axis)
tmax_tmin_plot <- ny_noaa_tidy %>%
  ggplot(aes(tmin, tmax)) +
  geom_hex()
# (ii) density plot
snwfall_plot <- ny_noaa_tidy %>%
  filter(0 < snow & snow < 100) %>%
  mutate(year = as.factor(year)) %>%
  ggplot(aes(x = snow, y = year)) +
  geom_density_ridges(alpha = .2)
  
tmax_tmin_plot + snwfall_plot
```

